---
title: "pathway_analysis_approach_2"
author: "Bhavika Kumar"
date: "2024-04-25"
output: html_document
editor_options: 
  chunk_output_type: console
---

Using rWiki Pathways

loading libraries
```{r warning=FALSE, message=FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
  library(DESeq2)
  library(ensembldb)
  library(AnnotationHub)
  library(GenomicFeatures)
  library(ggrepel)
  library(rWikiPathways)
  library(DOSE)
  library(GO.db)
  library(GSEABase)
  library(org.Hs.eg.db)
  library(clusterProfiler)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(stringr)
  library(RColorBrewer)
  library(RCy3)
})
```

loading data
KOPN sensitive vs resistant
```{r warning=FALSE, message=FALSE}
kopn_expr <- read.csv("d:/Users/Sujata Kumar/Desktop/Agin-Merin's Project/res_kopn8.csv", header = T)

# Preparing up and down regulated genes lists
up.genes <- kopn_expr[kopn_expr$log2FoldChange > 1 & kopn_expr$padj < 0.05, 1]
dn.genes <- kopn_expr[kopn_expr$log2FoldChange < -1 & kopn_expr$padj < 0.05, 1]
bkgd.genes <- kopn_expr[,1]
```

adding ENTREZ ID to the data
```{r warning=FALSE, message=FALSE}
# adding entrez id with up regulated gene dataset
up.genes.entrez <- clusterProfiler::bitr(up.genes, fromType = "ENSEMBL", toType = "ENTREZID", OrgDb = org.Hs.eg.db)
cat("\n\nWhich column contains my new Entrez IDs?\n")
head(up.genes.entrez)

keytypes(org.Hs.eg.db)

# adding entrez id with down regulated gene dataset
dn.genes.entrez <- clusterProfiler::bitr(dn.genes, fromType = "ENSEMBL", toType = "ENTREZID", OrgDb = org.Hs.eg.db)

# adding entrez id with rest of the gene dataset
bkgd.genes.entrez <- clusterProfiler::bitr(bkgd.genes, fromType = "ENSEMBL", toType = "ENTREZID", OrgDb = org.Hs.eg.db)
```

Gene Ontology- it aims to identify biological processes, cellular locations and molecular functions that are impacted in the condition studied. 
```{r warning=FALSE, message=FALSE}
egobp <- clusterProfiler::enrichGO(
  gene = up.genes.entrez[[2]],
  universe = bkgd.genes.entrez[[2]],
  OrgDb = org.Hs.eg.db, 
  ont = "BP",
  pAdjustMethod = "fdr",
  pvalueCutoff = 0.05,
  readable = T
)

# seeing the data first 10 rows
head(egobp, 10)

# plotting the results
barplot(egobp, showCategory = 20)
dotplot(egobp, showCategory = 20)
goplot(egobp)

ggplot(egobp[1:20], aes(x=reorder(Description, -pvalue), y=Count, fill=-p.adjust)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    scale_fill_continuous(low="blue", high="red") +
    labs(x = "", y = "", fill = "p.adjust") +
    theme(axis.text=element_text(size=11))
```

Enrichment analysis- creating an enrichment map, instead of nodes representing genes, nodes represent pathways or functions

GeneRatio = k/n

k is the overlap between your genes-of-interest and the geneset
n is the number of all unique genes-of-interest

BgRatio=M/N

M is the number of genes within each geneset
N is the number of all unique genes across all genesets (universe)

Term size refers to the number of genes associated with a given GO term or pathway
```{r warning=FALSE, message=FALSE}
egobp.results.df <- egobp@result
egobp.results.df$term.size <- gsub("/(\\d+)", "", egobp.results.df$BgRatio)
```

Filter for term size to keep only term size >=3, gene count >= 5 and subset
```{r warning=FALSE, message=FALSE}
egobp.results.df <- egobp.results.df[which(egobp.results.df[,'term.size'] >= 3 & egobp.results.df[,'Count'] >= 5),]
egobp.results.df <- egobp.results.df[c("ID", "Description", "pvalue", "qvalue", "geneID")]
```

Formatting gene list column
```{r warning=FALSE, message=FALSE}
egobp.results.df$geneID <- gsub("/", ",", egobp.results.df$geneID)
```

add column for phenotype
```{r warning=FALSE, message=FALSE}
egobp.results.df <- cbind(egobp.results.df, phenotype=1)
egobp.results.df <- egobp.results.df[, c(1, 2, 3, 4, 6, 5)]
```

change column headers
```{r warning=FALSE, message=FALSE}
colnames(egobp.results.df) <- c("Name","Description", "pvalue","qvalue","phenotype", "genes")
```

Saving the file
```{r warning=FALSE, message=FALSE}
egobp.results.filename <-file.path(getwd(),paste("clusterprofiler_cluster_enr_results.txt",sep="_"))
write.table(egobp.results.df,egobp.results.filename,col.name=TRUE,sep="\t",row.names=FALSE,quote=FALSE)
```

Pathway analysis
up regulated genes
```{r warning=FALSE, message=FALSE}
ewp.up <- clusterProfiler::enrichWP(
    up.genes.entrez[[2]],
    universe = bkgd.genes.entrez[[2]],
    organism = "Homo sapiens",
    pAdjustMethod = "fdr",
    pvalueCutoff = 0.1, #p.adjust cutoff; relaxed for demo purposes
)

head(ewp.up)
```

Adding gene symbols to the result object 
```{r warning=FALSE, message=FALSE}
ewp.up <- DOSE::setReadable(ewp.up, org.Hs.eg.db, keyType = "ENTREZID")
head(ewp.up)
```

Plotting...
```{r warning=FALSE, message=FALSE}
barplot(ewp.up, showCategory = 20)
dotplot(ewp.up, showCategory = 20)
```

down regulated genes
```{r warning=FALSE, message=FALSE}
ewp.dn <- enrichWP(
    dn.genes.entrez[[2]],
    #universe = bkgd.genes[[2]],  #hint: comment out to get any results for demo
    organism = "Homo sapiens",
    pAdjustMethod = "fdr",
    pvalueCutoff = 0.1, #p.adjust cutoff; relaxed for demo purposes
)

ewp.dn <- setReadable(ewp.dn, org.Hs.eg.db, keyType = "ENTREZID")

head(ewp.dn)
dotplot(ewp.dn, showCategory = 20)

enrichDF_down <- as.data.frame(ewp.dn)
enrichDF_up <- as.data.frame(ewp.up)
```

