---
title: "pathway_analysis_approach_1"
author: "Bhavika Kumar"
date: "2024-04-24"
output: html_document
editor_options: 
  chunk_output_type: console
---

Using DAVID annotation tool. Its a web based tool particularly for gene-enrichment analysis. 

loading libraries
```{r warning=FALSE, message=FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
  library(DESeq2)
  library(ensembldb)
  library(AnnotationHub)
  library(GenomicFeatures)
  library(ggrepel)
  library(pathview)
  library(pander)
  library(SPIA)
  library(org.Hs.eg.db)
})
```

Exploring the pathway below is not important

loading data
```{r warning=FALSE, message=FALSE}
kopn_degs <- read.table("d:/Users/Sujata Kumar/Desktop/Agin-Merin's Project/KOPN8_deg.tsv", header = T, sep = "")
```

selecting only the gene_id column as the david tool requires the list of gene id to be uploaded and saving it in a text file
```{r warning=FALSE, message=FALSE}
test_data <- kopn_degs %>% 
  dplyr::select(gene_id)

write.table(test_data, file = "d:/Users/Sujata Kumar/Desktop/Agin-Merin's Project/degs_names.txt", row.names = FALSE, quote = FALSE, col.names = FALSE)
```

Below shows the DEGs in the pathways and change in expression of each gene

```{r warning=FALSE, message=FALSE}
#Pathway ids for humans
data("paths.hsa")
pander(head(paths.hsa, n=5))

#checking gene.idtype argument possibiities
data("gene.idtype.list")
pander(gene.idtype.list)

#visuaising data
deseq.degs.logfc <- subset(merged_4, padj < 0.05, select = log2FoldChange)
pander(head(deseq.degs.logfc))

pathview(gene.data = deseq.degs.logfc, 
         pathway.id = "00020",
         species = "hsa")
```

###########################################################

Trying to look for topology of how genes interact with each other. In what pathway genes are involved and in what pathways the genes are inhibited. 

loading the data
```{r warning=FALSE, message=FALSE}
kopn_expr <- read.csv("d:/Users/Sujata Kumar/Desktop/Agin-Merin's Project/res_kopn8.csv", header = T)
reh_expr <- read.csv("d:/Users/Sujata Kumar/Desktop/Agin-Merin's Project/res_reh.csv", header = T)

# getting the ids vector
ids <- kopn_expr$row
ids <- reh_expr$row

# getting the entrez id in a different vector
entrez <- mapIds(org.Hs.eg.db, keys = ids, keytype = "ENSEMBL", column = "ENTREZID")

# converting entrez to data frame
entrez_df <- as.data.frame(entrez)

# converting row names to column
entrez_df$ENSEMBL <- row.names(entrez_df)

# changing the column name to ENSEMBL 
colnames(kopn_expr)[1] <- "ENSEMBL"
colnames(reh_expr)[1] <- "ENSEMBL"

# merging the data
merged_data <- merge(x=kopn_expr, y=entrez_df, all=T)
merged_data <- merge(x=reh_expr, y=entrez_df, all=T)

#loading merged 3 which has all the degs
merged_data[is.na(merged_data)] <- 0

# data subsetted which does not have any zeros
merged_subset <- merged_data[merged_data$entrez != 0, ]
```

Making the background genes and significant genes separately in differet vectors
```{r warning=FALSE, message=FALSE}
background_entrez <- merged_subset$entrez
sig_res_entrez <- merged_subset[which(merged_subset$padj < 0.05), ]
sig_entrez <- sig_res_entrez$log2FoldChange
names(sig_entrez) <- sig_res_entrez$entrez

head(sig_entrez)
```

Removing duplicates and NA values
```{r warning=FALSE, message=FALSE}
sig_entrez <- sig_entrez[!is.na(names(sig_entrez))] 

sig_entrez <- sig_entrez[!duplicated(names(sig_entrez))]

background_entrez <- merged_subset$entrez

background_entrez <- background_entrez[!is.na(background_entrez)]

background_entrez <- background_entrez[!duplicated(background_entrez)]
```

Running SPIA
```{r warning=FALSE, message=FALSE}
spia_result <- spia(de=sig_entrez, all = background_entrez, organism = "hsa")

head(spia_result)
```

Saving the spia result
```{r warning=FALSE, message=FALSE}
write.table(spia_result, file = "d:/Users/Sujata Kumar/Desktop/Agin-Merin's Project/spia_kopn_results.txt", row.names = FALSE)
write.table(spia_result, file = "d:/Users/Sujata Kumar/Desktop/Agin-Merin's Project/spia_reh_results.txt", row.names = FALSE)
```

###################################################################

Loading both SPIA data of reh and kopn
```{r warning=FALSE, message=FALSE}
spia_res_kopn <- read.table("d:/Users/Sujata Kumar/Desktop/Agin-Merin's Project/spia_results_kopn.txt", header = T)
spia_res_reh <- read.table("d:/Users/Sujata Kumar/Desktop/Agin-Merin's Project/spia_reh_results.txt", header = T)
```

Filtering cancer pathways from the datasets
```{r warning=FALSE, message=FALSE}
# from kopn datasets
filter_kopn <- spia_res_kopn %>% 
  dplyr::filter(Name=="Non-small cell lung cancer" 
                | Name=="Basal cell carcinoma" 
                | Name=="VEGF signaling pathway" 
                | Name=="Pathways in cancer" 
                | Name=="Prostate cancer" 
                | Name=="B cell receptor signaling pathway" 
                | Name=="Wnt signaling pathway" 
                | Name=="RNA degradation" 
                | Name=="Jak-STAT signaling pathway" 
                | Name=="Apoptosis" 
                | Name=="RNA transport" 
                | Name=="T cell receptor signaling pathway" 
                | Name=="NF-kappa B signaling pathway" 
                | Name=="Hedgehog signaling pathway" 
                | Name=="Pancreatic cancer" 
                | Name=="Small cell lung cancer" 
                | Name=="MAPK signaling pathway" 
                | Name=="Renal cell carcinoma" 
                | Name=="Thyroid cancer" 
                | Name=="mTOR signaling pathway" 
                | Name=="TGF-beta signaling pathway" 
                | Name=="PPAR signaling pathway" 
                | Name=="p53 signaling pathway" 
                | Name=="Endometrial cancer" 
                | Name=="Acute myeloid leukemia" 
                | Name=="Melanoma" 
                | Name=="Toll-like receptor signalling pathway" 
                | Name=="Chronic myeloiod leukemia" 
                | Name=="Notch signaling pathway" 
                | Name=="Colorectal cancer" 
                | Name=="Viral carcinogenesis" 
                | Name=="Transcriptional misregulation in cancer")

# from kopn datasets
filter_reh <- spia_res_reh %>% 
  dplyr::filter(Name=="Non-small cell lung cancer" 
                | Name=="Basal cell carcinoma" 
                | Name=="VEGF signaling pathway" 
                | Name=="Pathways in cancer" 
                | Name=="Prostate cancer" 
                | Name=="B cell receptor signaling pathway" 
                | Name=="Wnt signaling pathway" 
                | Name=="RNA degradation" 
                | Name=="Jak-STAT signaling pathway" 
                | Name=="Apoptosis" 
                | Name=="RNA transport" 
                | Name=="T cell receptor signaling pathway" 
                | Name=="NF-kappa B signaling pathway" 
                | Name=="Hedgehog signaling pathway" 
                | Name=="Pancreatic cancer" 
                | Name=="Small cell lung cancer" 
                | Name=="MAPK signaling pathway" 
                | Name=="Renal cell carcinoma" 
                | Name=="Thyroid cancer" 
                | Name=="mTOR signaling pathway" 
                | Name=="TGF-beta signaling pathway" 
                | Name=="PPAR signaling pathway" 
                | Name=="p53 signaling pathway" 
                | Name=="Endometrial cancer" 
                | Name=="Acute myeloid leukemia" 
                | Name=="Melanoma" 
                | Name=="Toll-like receptor signalling pathway" 
                | Name=="Chronic myeloiod leukemia" 
                | Name=="Notch signaling pathway" 
                | Name=="Colorectal cancer" 
                | Name=="Viral carcinogenesis" 
                | Name=="Transcriptional misregulation in cancer")
```

selecting name of the pathway, status and p-value (column pG) 
```{r warning=FALSE, message=FALSE}
# kopn data
select_kopn <- filter_kopn %>% 
  dplyr::select(Name, pG, Status, pGFdr, pPERT)

# reh data
select_reh <- filter_reh %>% 
  dplyr::select(Name, pG, Status, pGFdr, pPERT)
```

adding column cell line in the data
```{r warning=FALSE, message=FALSE}
select_kopn$cell_line <- "kopn"
select_reh$cell_line <- "reh"
```

merging the datasets
```{r warning=FALSE, message=FALSE}
merged_data <- merge(x=select_kopn, y=select_reh, by="Name")
merged_2 <- merged_data %>% 
  dplyr::select(Name, Status.x, Status.y)
colnames(merged_2)[2] <- "Status_KOPN"
colnames(merged_2)[3] <- "Status_REH"

df <- melt(merged_2, id.vars = "Name", variable.name = "Status")

ggplot(df, aes(value, Name)) + geom_point(aes(colour = Status))

ggplot(merged_2) + 
  geom_point(aes(x= Status_KOPN, y=Name, colour = Status_KOPN))+
  geom_point(aes(x=Status_REH, y=Name, colour = Status_REH))


theme_bw() + labs(x="Status", y="Names of the pathways", title = "KOPN")
```

plot to see which pathways are inhibited and activated in which cell line
```{r warning=FALSE, message=FALSE}
ggplot(merged_data, aes(x=Status.x, y=Name, colour = Status.x)) + geom_point() + theme_bw() + labs(x="Status", y="Names of the pathways", title = "KOPN")

ggplot(merged_data, aes(x=Status.y, y=Name, colour = Status.y)) + geom_point() + theme_bw() + labs(x="Status", y="Names of the pathways", title = "REH") 

ggplot(select_kopn, aes(x=pGFdr, y=Name)) + geom_point() + theme_bw() + labs(x="pGFDR", y="Names of the pathways", title = "KOPN8") 

ggplot(select_reh, aes(x=pGFdr, y=Name)) + geom_point() + theme_bw() + labs(x="pGFDR", y="Names of the pathways", title = "REH") 
```

Significant pathways
selecting pathways that are less than 0.05
```{r warning=FALSE, message=FALSE}
sig_kopn <- select_kopn %>% 
  dplyr::filter(pG < 0.05)
sig_reh <- select_reh %>% 
  dplyr::filter(pG < 0.05)
```

Taking the value of pPERT as less than 0.7
```{r warning=FALSE, message=FALSE}
less_than_kopn <- select_kopn %>%
  dplyr::filter(pPERT < 0.7)
less_than_reh <- select_reh %>% 
  dplyr::filter(pPERT < 0.7)

ggplot(less_than_kopn, aes(x=pPERT, y=Name)) + geom_point(aes(colour = Status)) + theme_bw() + labs(x="pPERT", y="Names of the pathways", title = "KOPN") 

ggplot(less_than_reh, aes(x=pPERT, y=Name)) + geom_point(aes(colour = Status)) + theme_bw() + labs(x="pPERT", y="Names of the pathways", title = "REH") 
```

